\documentclass[a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{twocolceurws}


\title{Damegender: Writing and Comparing Gender Detection Tools}

\author{
David Arroyo Menéndez \\ GSYC \\ Universidad Rey Juan Carlos I \\ d.arroyome@alumnos.urjc.es
\and
Jesús González Barahona \\ GSYC \\ Universidad Rey Juan Carlos I \\ jgb@gsyc.es
}

\institution{}


\begin{document}
\maketitle

\begin{abstract}
The variable sex (male or female) is one of most used variables for
any study in sociology, but this variable can be hidden in Internet
communities. The gender detection from a name is an important problem
in Natural Language Processing to decide if a string labeled as name
is classified as male or female. An engineer will find useful
make gender detection from a name retrieving information from social
networks, mailing lists, instant messaging, software repositories,
papers, etc. To achieve gender equality and empower all women and
girls is a goal in sustanaible development in United Nations, so to
measure the gender gap is a previous step to find solutions to reduce
it.

Nowadays, there are several Application Programming Interfaces to
guess gender from a name. This kind of software has the database
based on propietary databases and the software is not free, so some
scientific works are difficult to reproduce.

In this paper, we are envisioning how to solve these problems,
offering a solution with a free license and open data names from
official census useful to replace, use and/or compare these apis with
very good results. This tool provides Machine Learning to predict
strings, that's useful to guess diminutives or nicknames.
\end{abstract}


\section{Introduction}

There are different ways to detect gender from a person name and
perhaps a surname: census, wikipedia, self-references in trust
websites, ... The most common way to detect gender from a name is the
Application Programming Interfaces with a good popularity, for
example, genderapi, namsor, genderize, ...
~\cite{10.7717/peerj-cs.156}

The problems addressed are:
\begin{itemize}
\item Evaluate quality/price with different commercial solutions.
\item Think about solutions using free licenses.
\item Treatment with names without census, for example, nicknames,
  diminutives, ...
\item Massive gender detection from Internet, for example, mailing
  lists, software repositories, ...
\end{itemize}

In this paper, these problems are faced writing a Python solution for:
\begin{itemize}
\item To evaluate quality of different solutions applying metrics
  suggested by ~\cite{10.7717/peerj-cs.156}
\item To understand the current technology in detail, I have developed
  a tool guessing gender from a name giving support to Spanish and
  English from the open data census provides by the states.
\item To fix the problem with nicknames and diminutives, we have
  developed a machine learning solution to strings not found in the
  census dataset.
\item To do proof-of-concept tests applying Perceval to detect
  gender in mailing lists and software repositories.
\end{itemize}

In Section 2, we explain the current solutions to the problems. In
Section 3, we present the results evaluating the current Application
Programming Interfaces with our software. In Section 4, we discuss
attempts and problems releasing with a free license a gender detection
from name program. In Section 5, we discuss how to obtain Open Datasets
counting names and gender. In Section 6, we describe our
machine learning solution. In Section 7, we describe general
implementation details. Finally, in Section 8 we summarize our
findings, and describe extensions to the work that we are currently
exploring.

\section*{3. Application Programming Interfaces}

\subsection*{Market}

We have reproduced to ~\cite{10.7717/peerj-cs.156} and updated on
27/06/2019 and we are showing the results in
~\ref{table:featuresAndGenderDetectionToolsByName}.

\begin{longtable}[]{@{}lllllll@{}}
\toprule
Feature & Gender API & genderguesser & genderize.io & NameAPI & NamSor & Damegender\tabularnewline
\midrule
\endhead
Database size & 431*10^6 & 48.528 & 114*10^6 & 1.428.345 & 4407*10^6  & 197.271\footnote{This cipher is the sum of names in United States of America, United Kingdom, Uruguay and Spain} \tabularnewline
Regular data updates & yes & no & yes & yes & yes & yes, dev\tabularnewline
Unstructured strings & yes & no & no & yes & no & yes\tabularnewline
Handles surnames & yes & no & no & yes & yes & yes\tabularnewline
Non-Latin alphabets & partially & no & partially & yes & yes & no\tabularnewline
Geo-localization & yes & no & no & yes & yes & no\tabularnewline
Exists locale & yes & yes & yes & yes & yes & yes\tabularnewline
Assingment type & probabilistic & binary & probabilistic & probabilistic & probabilistic & prob\tabularnewline
Free params & total, prob & gender & total, prob & confidence & scale & total, prob\tabularnewline
Guessing with ML & no & no & no & no & no & yes\tabularnewline
Free license & no & yes & no & no & no & yes\tabularnewline
API & yes & no & yes & yes & yes & future\tabularnewline
free requests limited & yes (200) & unlimited & yes (1000) & yes & yes & unlimited\tabularnewline
\bottomrule
\caption{Features and gender detection tools by name}
\label{table:featuresAndGenderDetectionToolsByName}
\end{longtable}

All solutions have increased the database size from
~\cite{10.7717/peerj-cs.156}. NameAPI and GenderAPI is reaching more
features. The tools with a free license have not many features, so for
now that will not be the trend in many situations. Today, one good
solution quality and price is Namsor, which provides unlimited names
through an Application Programming Interface with many features in the
task to detect gender from the name.

\subsection*{Reproducing accuracies and confusion matrix}

~\cite{10.7717/peerj-cs.156} explains different ways to determine
gender from a name by humans and it gives 7000 names applying these
methods. In this dataset the gender is classified as male, female or
unknown. We have used this dataset, but only male and female to these
experiments. We are showing the results in the next table:

\begin{longtable}[]{@{}lllllll@{}}
  \toprule
  API & Accuracy & Precision & F1score & Recall\tabularnewline
\midrule
\endhead
Genderapi & 0.9687686966482124 & 0.9717050018254838 & 0.9637877964874163 & 1.0\tabularnewline
Genderize & 0.926775 & 0.9761303240374678 & 0.9655113956503119 & 1.0\tabularnewline
Damegender (SVC)\footnote{SVC is the acronym of Support Vector Classification, the Machine Learning algorithm that Damegender was using with this results} & 0.8791969539633091 & 0.9718767935718385 & 0.9718767935718385 & 1.0\tabularnewline
Namsor & 0.8672551055728626 & 0.9730097087378641 & 0.9236866359447006 & 1.0\tabularnewline
Nameapi & 0.8301886792452831 & 0.97420272191753 & 0.9054181612233341 & 1.0\tabularnewline
Gender Guesser & 0.7743554248139817 & 0.9848151408450704 & 0.8715900233826968 & 1.0\tabularnewline
\bottomrule
\caption{Different accuracies measures}
\label{table:DifferentAccuraciesMeasures}
\end{longtable}

In ~\ref{table:DifferentAccuraciesMeasures} Genderapi and Genderize
are obtaining the best results, although all solutions is reaching
results better than 0.8 except Gender Guesser.

\begin{longtable}[]{@{}lllll@{}}
  \toprule
  APIs          & gender & male & female & undefined \tabularnewline
\midrule
\endhead
Genderapi         & male    & 3589 & 155  &  67 \tabularnewline
                  & female  & 211  & 1734 &  23 \tabularnewline
Damegender (SVC)\footnotemark[1]  & male    & 3663 & 147  &   0 \tabularnewline
                  & female  & 551  & 1497 &   0 \tabularnewline
Genderguesser     & male    & 3326 &  139 & 346 \tabularnewline
                  & female  & 78   & 1686 & 204 \tabularnewline
Namsor            & male    & 3325 & 139  & 346 \tabularnewline
                  & female  & 78   & 1686 & 204 \tabularnewline
Genderize         & male    & 3157 & 242  & 412 \tabularnewline
                  & female  & 75   & 1742 & 151 \tabularnewline
Nameapi           & male    & 2627 & 674  & 507 \tabularnewline
                  & female  & 667  & 1061 & 240 \tabularnewline 
\bottomrule
\caption{Confusion matrix tables by APIs}
\label{table:ConfusionMatrixTables}
\end{longtable}

With Damegender has been done a comparison about confusion matrix
tables depending the API (see ~\ref{table:ConfusionMatrixTables}). If
we compare these results with the results obtained in
~\cite{10.7717/peerj-cs.156}, we can understand that the results are
similar.

Genderapi has similar results, but it is being improved the undefined
results. In Genderguesser is we are obtaining different results and it
is strange, because the software has not modified from some years
ago. In Genderize we are obtaining the same results. In Nameapi the
guessed results is changing from male to female with more errors. In
Namsor the results is so similar. Damegender is not guessing undefined
because we predict with machine learning if the string is not in the
database.

The most important tools Namsor, Genderapi and Genderize are improving the
accuracies with respect the previous comparison.

\begin{longtable}[]{@{}lllll@{}}
\toprule
API & error code & error code without na & na coded & error gender bias\tabularnewline
\midrule
\endhead
Damegender (SVC)\footnotemark[1] & 0.121 & 0.121 & 0.0 & -0.07\tabularnewline
GenderApi & 0.167 & 0.167 & 0.0 & -0.167\tabularnewline
Gender Guesser & 0.225 & 0.027 & 0.204 & 0.003\tabularnewline
Genderize & 0.276 & 0.261 & 0.0204 & -0.0084 \tabularnewline 
Namsor & 0.332 & 0.262 & 0.095 & 0.01 \tabularnewline
Nameapi & 0.361 & 0.267 & 0.129 & 0.001 \tabularnewline
\bottomrule
\caption{APIs and Errors}
\label{table:ApisAndErrors}
\end{longtable}

In the table it is possible to observe a high index of errors in
Nameapi and Namsor and a low index of errors in GenderApi
and Damegender.


\section*{Datasets}

We can divide the next options choosing a dataset: (1) a census
published with a free license (open census way), (2) a dataset done by
scientist with a paper in a magazine (scientific way), (3) a dataset
released with a free license in a free software package (free software
way), (4) a dataset retrieved from commercial Application Programming
Interfaces (commercial api way).

\begin{verbatim}
$ python3 main.py David --total="ine"
David gender is male
363559  males for David from INE.es
0 females for David from INE.es
\end{verbatim}

In Damegender, we are including Open Data census about names and
gender, such as INE.es or USA and United Kingdom (births and dies). We
want datasets provided by the software package to increment the speed
retrieving data.

From the user final point of view, the value of using Open Data is
give a good explanation when we are asking about the gender from a
name (number of males and females using a specific name in a country)
versus a probability created by the way explained in
~\cite{10.7717/peerj-cs.156} or similar.

From the scientific point of view, the value of using Open Data is to
allow that the experiment can be reviewed by peers on an automatic and
legal way (using proprietary data the reviewer should request it
separately to make the review).

A second approach is to build the dataset reviewing the names in
scientific personal sites, Wikipedia,
... ~\cite{10.7717/peerj-cs.156}. This approach is valid, but it
consumes many time and efforts, although could be useful if there not
a legal way to build the dataset.

A third approach is using a dataset from a popular free software
solution. For instance, Natural Language Tool Kit is providing 8000
labeled english names. The classification is male or female. The
problem again is about don't retrieve data with the social science
quality of National Statistics Institutes. Another example is Gender
Guesser a good dataset for international names with different
categories to define the probability. This approach is similar to use
a dataset released with a paper in a journal, the advantage is to
understand and add new names with a solid criteria accepted by the
scientific community.

We are using the census approach as base of truth to distinguish if a
name is male or female in a geographical area. Generally, a name has a
strong weight to determine if it's a male or a female on this way, for
instance, David is registered 365196 times as male and 0 times as
female in Spain National Institute of Statistics.

Many countries don't provide Open Data census about gender and names,
but we envisioned build a Dataset about names and gender free and
universal working from Gender Guesser dataset and Wikidata as
solution. Perhaps, to complete this work we need automate humans
process described in ~\cite{10.7717/peerj-cs.156}.

The last approach is based on to trust on commercial solutions, such
as we trust on search engines to make searches in Internet (black
box).  In Damegender we can download json files from main commercial
Application Programming Interfaces (API) solutions (genderapi,
genderize, namsor, nameapi, ...). One user can build proprietary
datasets on this way using an average weighted by the precision or
accuracy of each Application Programming Interface measured with
Damegender with an open dataset as base of truth.

\section{Machine Learning}

We have developed a script infofeatures.py with our datasets to
analyze data about features. The datasets used in this experiment was
retrieved from official datasets from national statistical
institutions in Spain, Uruguay, United Kingdom, USA. The features used
are: first letter, last letter, a, b, c, d, e, f, g, h, i, j, k, l, m,
n, o, p, q, r, s, t, u, v, w, x, y, z, vocals, consonants, first
letter, first letter vocal, last letter vocal, last letter consonant,
last letter a. The choosing of features was verified with Principal
Component Analysis.

Take a look to the most informative features results with the
different datasets:
\begin{longtable}[]{@{}llllllll@{}}
  \toprule
Dataset & Char A & Last Char A & Last Char O & Last Char Consonant & Last Char Vocal & First Char Consonant & First Char Vocal & \tabularnewline
  \midrule
  \endhead
 Uruguay (F) \footnote{F is for females and M is for males} &    0.816 &         0.456 &         0.007 &                 0.287 &             0.712 &                  0.823 &              0.177 & \tabularnewline
 Uruguay (M) \footnotemark[3] &    0.643 &         0.249 &         0.062 &                 0.766 &             0.234 &                  0.771 &              0.228 & \tabularnewline
 Spain (F) \footnotemark[3]  &    0.922 &         0.588 &          0.03 &                 0.271 &             0.728 &                  0.772 &              0.228 & \tabularnewline
 Spain (M) \footnotemark[3]  &    0.818 &          0.03 &         0.268 &                 0.569 &              0.43 &                  0.763 &              0.236 & \tabularnewline
 UK (F) \footnotemark[3]     &    0.825 &         0.374 &         0.013 &                 0.322 &             0.674 &                  0.765 &              0.235 & \tabularnewline
 UK (M) \footnotemark[3]     &    0.716 &         0.036 &         0.039 &                  0.78 &             0.218 &                  0.799 &                0.2 & \tabularnewline
 USA (F) \footnotemark[3]    &    0.816 &         0.456 &         0.007 &                 0.287 &             0.712 &                  0.823 &              0.177 & \tabularnewline
 USA (M) \footnotemark[3]    &    0.643 &          0.02 &         0.061 &                 0.765 &             0.234 &                   0.84 &              0.159 & \tabularnewline
\bottomrule
\caption{Informative Features in Different Countries}
\label{table:InfoFeatures}
\end{longtable}

The countries where the main language is spanish (Uruguay + Spain) and
english (USA + United Kingdom + Australia) where is having very
similar variation with the features chosen between males and females
with these datasets (remember that this datasets are being extracted
from official statistics provided by the states). In Canada, a country
french centric has different rules with this features.

The letter a is varying 0.2 from males to females in (USA and Uruguay)
and 0.1 from males to females (United Kingdom, Australia and
Spain). The last letter a is varying 0.5 from males to females in
(Australia, Spain) around 0.4 in (USA, United Kingdom) and 0.2 in
Uruguay. The last letter o from females to males is varying 0.2 in
(Spain, Australia) and is equal in (Uruguay, USA, United Kingdom). For
the last letter consonant all countries is giving the result that is
for males, with results from 0.2 to 0.5: Uruguay and USA (0.5), United
Kingdom (0.4), Australia and Spain (0.3). So last letter vocal is
reverse tha last letter consonant. First letter consonant or first
letter vocal is a non significative feature due to so similar results
in english and spanish.

The success with the different algorithms is showed in the next table:

\begin{longtable}[]{@{}lllll@{}}
  \toprule
  Machine Learning Algorithm & Accuracy & Precision & F1score & Recall \tabularnewline
  \midrule
  \endhead

 Support Vector Machines             &    0.879 &     0.972 &   0.972 &    1.0 & \tabularnewline
 Random Forest                       &    0.862 &     0.902 &   0.902 &    1.0 & \tabularnewline
 NLTK (Bayes)                        &    0.862 &     0.902 &   0.902 &    1.0 & \tabularnewline
 Multinomial Navie Bayes             &    0.782 &     0.791 &   0.791 &    1.0 & \tabularnewline
 Tree                                &    0.764 &     0.821 &   0.796 &    1.0 & \tabularnewline
 Stochastic Gradient Distribution    &    0.709 &     0.943 &   0.815 &    1.0 & \tabularnewline
 Gaussian Naive Bayes                &    0.709 &     0.968 &   0.887 &    1.0 & \tabularnewline
 Bernoulli Naive Bayes               &    0.699 &     0.965 &   0.816 &    1.0 & \tabularnewline
 AdaBoost                            &    0.698 &     0.965 &   0.815 &    1.0 & \tabularnewline
 Multi Layer Perceptron              &    0.677 &     0.819 &   0.755 &    1.0 & \tabularnewline

\bottomrule
\caption{Machine Learning Algorithms and accuracies measures}
\label{table:MLAccuracies}
\end{longtable}

The results in ~\ref{table:MLAccuracies} shows that using algorithms
as Support Vector Machines or Random Forest against a scientific
dataset created by independent researchers where it is possible to
reach results similar to another commercial solutions about gender
detection tools. Our classifier is binary (only male and female).

We were doing this experiment with NLTK and INE datasets with
accuracies reaching accuracies until 0.745. So it makes sense expect
better results in random datasets augmenting languages and
countries. Due to our solution is not providing arabic or chinesse
alphabets, yet.

So, it's possible infer that Damegender provides a good solution for
nicknames, diminutives, or similar.


\section{State of Art}

\subsection{Comparing Commercial Solutions}

A standard commercial Application Programming Interface (API) can
guess the gender for a single name or a list of names (from a CSV file
or an API call). To express geolocalization you can give surnames, a
country ISO code, or a language. Generally, you can give a probability
and a counter associated to a name and gender in a certain population.

~\cite{10.7717/peerj-cs.156} are proposing a good metrics set to
classify these commercial Application Programming Interfaces (features,
measuring errors and success, ...). The features observed are:
Database size (January 2018), Regular data updates, Handles
unstructured full name strings, Handles surnames, Handles non-Latin
alphabets, Implicit geo-localization, Assignment type, Free
parameters, Open source, Application Programming Interface, Monthly
free requests, Monthly subscription cost (100,000 requests/month)).

In the commercial tools is being used different ways to express
probability (confidence, scale, accuracy, precision, recall, ...). 

\subsubsection{Datasets}

In ~\cite{berners2001semantic} a world was envisioned where public
structured data could be interconnected with software agents to
process these data, perhaps only recovering information, but mixed
with distributed artificial intelligence would give a big jump to the
semantic richness to the web.

~\cite{janssen2012benefits} shows serious profits for the states
adopting Open Data in three categories (1) political and social, (2)
economical, (3) operational and technical. So, Open Data is a
breakthrough towards the Semantic Web.

We can find Open Data about names and gender in census of citizens in
states and commercial solutions. Free software packages such as
~\cite{krawetz2006gender} or ~\cite{loper2002nltk} is providing good
datasets about names and gender. So, Damegender incorporates different
lists of names from free software solutions wrote before (Natural
Language ToolKit, Gender Guesser, ...) and from Open Data census
(United Kingdom, USA, Spain, Uruguay, ...). 

Wikidata ~\cite{42240} provides a semantic and open database about
Wikipedia allowing retrieve information with Sparql, such as names and
gender.

~\cite{10.7717/peerj-cs.156} describes different ways to build a
dataset on hand looking for names in papers, scientific websites,
wikipedia, biographies, photos, ...)


\subsection*{Free Software}

Before Damegender, only ~\cite{krawetz2006gender} was competing as
Free Software solution with the main commercial Application
Programming Interfaces about gender detection from the name. The best
contribution is the dataset containing 48528 names with a good
classification by countries.

\subsection*{More software about gender}

In some studies, for example, about Twitter or Github, some people can
choose between different ways to detect gender (not only names). So,
we can find gender detection tools from faces in images
(~\cite{ranjan2017hyperface}), from hand written
(~\cite{liwicki2011automatic}), or from speeches
(~\cite{koppel2002automatically}).

\subsection*{Massive Gender Detection}

There are good studies measuring gender in Internet. Some studies are
about gender gap in general (~\cite{robles2014floss},
~\cite{holman2018gender}, ~\cite{dollar1999gender}), Twitter
(~\cite{burger2011discriminating}, ~\cite{mislove2011understanding})
Stackoverflow (~\cite{vasilescu2012gender}), Wikipedia
(~\cite{antin2011gender}, ~\cite{hill2013wikipedia}), Github
(~\cite{vasilescu2015gender}) ...

\section*{Implementation}

We have chosen Python free software tools with a good scientific
impact. Natural Language Toolkit for Natural Language Processing
~\cite{loper2002nltk}. Scikit for Machine Learning
~\cite{pedregosa2011scikit}. Numpy for Numerical Computation
~\cite{van2011numpy}. Matplotlib to visualize results
~\cite{hunter2007matplotlib}. And Perceval ~\cite{duenas2018perceval}
to retrieve information in mailing lists and repositories.

The current result is a Python package contributed to pip to be used
from the console.

The software is using an oriented to objects design with unit testing
for classes and methods using nose and unit testing for Python
commands using Bash. 

A summary of current features in the software are:

\begin{itemize}[noitemsep]
\item To deduce the gender about a name in Spanish or English (current
  status) from open census in local.
\item To decide about males and females in strings using different
  machine learning algorithms.
\item To use the main solutions in gender detection (genderize, genderapi,
namsor, nameapi and gender guesser) from a command.
\item To research about why a name is related to males or females with
  statistics. We provide Python commands about study and compare
  gender solutions with: confusion matrix, accuracies, error
  measures. And to decide about features: statistical feature weight,
  principal component analysis, ...
\item To determine gender gap in free software repositories or mailing
  lists (proof of concept)
\end{itemize}


\section*{Conclusions}

The market of gender detection tools is dominated by companies based
on payment services through Application Programming Interfaces with
good results. This market could be modified due to Free Software
tools and Open Data giving more explicative results for the user.

Although machine learning techniques are not new in this field, we
are giving an approach to guess strings not found in a dataset that
currently is classified as unknown and the humans trend to think in
gender terms many strings calling it as nicknames or diminutives.

These previous advances in computer science could be giving support to
study the gender gap in repositories and mailing lists. Another future
work is to create a free and universal dataset with support for all
languages and cultures.

\end{document}

%\bibliographystyle{alpha} 
%% \bibliography{satbibtex}

%% \bibitem[Com79]{Comer-btree}
%% D.~Comer.
%% \newblock The ubiquitous b-tree.
%% \newblock {\em Computing Surveys}, 11(2):121--137, June 1979.

%% \bibitem[Knu73]{Knuth-vol3}
%% D.~E. Knuth.
%% \newblock {\em The Art of Computer Programming -- Volume 3 / Sorting and
%%   Searching}.
%% \newblock Addison-Wesley, 1973.


%% \bibliographystyle{alpha} 
%% \bibliography{bibtex}
%inline the .bbl file directly for mailing to authors.


 
